Counting evolutions: a simple foundation for thermodynamics

### 2021-01-05 Articles/2019-Entropy/Source/thermodynamics.tex
First draft for thermodynamics

\title{Counting evolutions: a simple foundation for thermodynamics}
\author{Gabriele Carcassi, Christine A. Aidala, et al \\ University of Michigan}

@@

This draft summarizes the status of our work on a new approach to the foundations of thermodynamics and is intended to gather early feedback. Rather then derive thermodynamics bottom-up from a particular microscopic model, we want to derive it top-down from a set of requirements on the dynamics of the system. The results we find are, then, independent on the microscopic dynamics in the sense that they can in principle be realized in different ways. This is both more in line with the aims of the founders of thermodynamics and justifies why similar concepts have been successful in fields beyond physics. Moreover, thermodynamics is found as a specialization of a more general model, which we believe can be used as a common foundation for the different types mechanics.

@@

There are many technical details that are required to make the above characterization formally rigorous, mostly due to the proper handling of infinities. However, those do not really change the overall intuition, which we believe is the main strength of the approach. Therefore, in this paper, we will only give a conceptual overview leaving the formal details to other. This will leave the key physical insight is more directly accessible, and also result in a narrative that can be used directly in an introductory course in thermodynamics.

@@

Conversely, we say a process is \textbf{reversible} if the state at one time is enough to reconstruct the state at all past times. We can see an example of such a process in figure \ref{fig_reversibility}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all past states: evolutions can diverge but cannot merge. The count of evolutions per state over time can never increase: $
\mu(x(t)) \geq \mu(x(t + \Delta t))$.

@@

Note that in some literature the ability to reconstruct the past is called retrodictability and does not coincide with thermodynamic reversibility (the ability to undo a change). This distinction will be made precise and clear in the full framework. At this point we want to make it clear that our notion of reversibility will indeed correctly recover thermodynamic reversibility. It will do so without having to assume the existence of a reverse process or a symmetry under time-reversal.

@@

\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide5.png}
	\caption{Determinism and reversibility. Evolutions cannot merge nor split. The number of evolutions per state cannot increase.}\label{fig_detrev}
\end{figure}

To finish, we focus on deterministic processes with equilibria. We can see an example of such a process in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left. In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

Note that the idea of equilibrium as ``nothing changing'' used here is an oversimplification, which does not even work relativistically. Yet, it suffices for now as it captures the intuition correctly.

@@

Note that we have found a quantity, $\mu(x(t))$, that can never decrease in a deterministic process, stays the same if the process is reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, expect for one thing we still need to understand: how it combines under system composition.

Suppose we now have two systems, with their respective state space $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\{x(t), y(t)\}$ of the composite system. In general, the state of one system may affect the other system, in which case the evolution of the composite cannot be separated. If the systems are independent, however, the evolution of one system will not influence the other. That is, for each possible evolution $x(t)$ of the first system and $y(t)$ of the second system, there will be a possible evolution $\{x(t), y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under deterministic process with equilibria. Which means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems. As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$, which we will address in the full framework.

In general, the process entropy will depend on the process, not just on the state. However, a unique value for process entropy can be assigned to states of equilibrium. Suppose, in fact, that the state $x$ is an equilibrium for two different processes $A$ and $B$. Suppose that $A$ is at equilibrium $x$ and we switch to process $B$. Since $x$ is also an equilibrium for $B$, every evolution of $A$ that ended in $x$ will correspond to an evolution in $B$ that ends in $x$. By symmetry, the converse is also true. Therefore the process entropy of an equilibrium does not depend on the process, only on the equilibrium state. This allows us to assign a unique process entropy to the state. If we have set of states that are equilibria, then, we call \textbf{state entropy} $S(x)$ the process entropy that the system reaches at equilibrium.

To summarize, we found, under most general conditions, a quantity called process entropy, defined as the logarithm of the number of evolutions, that cannot decrease under deterministic processes, cannot increase under reverersible processes, is maximized at equilibrium and is additive for independent systems. Moreover, if $X$ consists of equilibria, there exists an \textbf{equation of state} $S : X \to \mathbb{R}$ which returns the state entropy, which is the process entropy of the equilibrium.

@@

We now have all the basic tools we need to reconstruct thermodynamics, which we characterize as the study of equilibria of composite systems. This is not done in general, but under additional assumptions and definitions that tell us how the equilibrum of the whole relates to the equilibria of the parts. As such, the laws we find are not necessarily universal in nature: they apply to systems that satisfy (or can be assumed to approximately satisfy) the prerequisites.

We say a system is in \textbf{thermodynamic equilibrium} if:
\begin{itemize}
	\item it is at equilibrium
	\item all its parts are in thermodynamic equilibrium
	\item if another system joined with a part of the system in thermodynamic equilibrium is in thermodynamic equilibrium, then the whole is in thermodynamic equilibrium.
\end{itemize}
We say two or more systems are in thermodynamic equilibrium if their composite is in thermodynamic equilibrium. A \textbf{thermodynamic process} is a process with equilibria where the final state is a thermodynamic equilibrium. We again stress that we are not saying that two systems, in general, always allow for thermodynamic equilibria.

@@

The characterization of thermodynamic equilibrium we give is manifestly a part-whole relationship. It extends the notion of equilibrium to a composite system so that the equilibrium of the whole and the equilibrium of its parts are essentially the same equilibrium. We find that this kind of part-whole relationships are a good starting point not only for thermodynamics, but also for the other branches of physics. This should not come as a surprise as reductionism (i.e. explaining the whole through its parts) plays an important role in science.

We define a \textbf{thermodynamic system} one whose equilibria can be identified by a set of extensive quantities (i.e. additive under system composition). Moreover, one such quantity is called internal energy which has the property of remaining conserved under any process. We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities. For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities. This is simply the equation of state in entropic form, which is the starting point of Gibbs formulation of thermodynamics. [CITE] We can therefore define the relationships
\begin{align}
	\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ dimensionless..

Note, again, that the system is identified by quantities that have a clear rule under system composition. We suspect that the existence of extensive quantities can be derived from the composability of thermodynamic equilibria. In fact, we typically construct measurement scales of quantities (such as energy, mass, volume, distance, ...) by preparing unit samples and then adding or dividing them. So it may be possible that, once we assume the ability to combine systems into bigger systems, we will consequently be able to construct extensive quantities without extra assumptions.\footnote{XXX: any work related to this?} Nevertheless, we leave these considerations for later works.

@@

To derive the first and the second law we need to give a definition for work and heat. We do so by defining two special thermodynamic systems. We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if its equation of state can be sufficiently approximated by $S = \beta U = \frac{U}{k_B T}$. Internal energy is the only state variable and the only way the entropy of the system can change. We call \textbf{heat} the energy exchanged by a thermal reservoir and we note $Q = - \Delta U$ as the energy lost by the reservoir during a thermodynamic process.

We say $M$ is a mechanical system if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S(U, x^i) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU &= k_B T dS + X_i dx^i = X_i dx^i.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U$ as the energy gained by the system during a thermodynamic process.

These two types of systems allow us to study the two ``orthogonal'' directions a change of energy can have: an energy change in the direction of constant entropy and one in the direction of maximal energy change. These definitions makes this manifestly clear. Also note that $\frac{\partial S_M}{\partial U} = 0 = \beta_M = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.

We now consider a system composed of a generic system $A$, a thermal reservoir $R$ and a mechanical system $M$. During a thermodynamic process, because the total energy of the system must be conserved, we have:
\begin{equation}
\begin{aligned}
&\Delta U_A + \Delta U_R + \Delta U_M = 0 \\
&\Delta U_A - Q + W = 0 \\
&\Delta U_A = Q - W.
\end{aligned}
\end{equation}
This recovers the first law.

@@

As state entropy counts the possible evolutions of the system at equilibrium, the lowest possible count $\mu$ is one, which is achieved if the equilibrium can be reached only by one starting condition. In this case, $S=\log(\mu)=0$. This tells us that state entropy is non-negative and therefore is bounded from below. We now have to show that all thermodynamic systems must have a lowest energy state and that it must be a local minima for entropy.

First of all, the existence of unique thermodynamic equilibria forces $S(U)$ to be strictly concave (i.e. negative second derivative). This means that, at least in some region, $S(U)$ must either be a decreasing or increasing function. But since $S$ cannot be negative, $S(U)$ cannot be defined on the whole domain. Therefore $U$ must be bounded at least from above or below. Consequently, the minimum for entropy is reached either at the lowest or highest energy state. Now we show that all thermodynamic systems must have either a lowest energy state or a highest energy state. Suppose, in fact, we had two system, $A$ has no maximum energy while $B$ has no minimum energy. Since $S_A(U_A)$ is concave, it must be a monotonically increasing function of $U_A$. This means that $\beta_A$ is always positive. Conversely, $S_B(U_B)$ must be monotonically decreasing, which means $\beta_B$ is always negative. If we put the two system in contact, the energy of $B$ will keep decreasing and the energy of $A$ will keep increasing and no equilibrium will ever be reached. Therefore all thermodynamic systems either have a minimum or a maximum energy. At this point the difference is a sign convention, so we can assume all thermodynamic systems have a minimum energy.

@@

\section{Discussion}

We have shown that thermodynamics can be conceptually rederived by adding few other assumptions/definitions from a general characterization of states, processes and equilibria. Can this approach also offer new insights on classical mechanics, quantum mechanics and statistical mechanics that may lead to a more unified understanding these somewhat disconnected disciplines?

The notion of equilibria would be a place to start. We defined, as it is customary, the state space of a thermodynamic system as consisting only of states of equilibria. We also defined a mechanical system as one for which $S$ is the same for all states. But we said the state entropy is uniquely defined only for state of equilibria. This implicitly assumed that all states of a mechanical systems are equilibria. Is this always true? And in what sense is this true?

Let us first consider the classical phase space of a particle. Qualitatively, we can imagine saying that a particle at rest is in equilibrium. A particle in uniform motion, though, does not seem in equilibrium: the distance is constantly increasing. Yet, there exists a frame in which it is at rest. So, in this a sense, all free particles are at equilibrium. Going one step further, in the presence of gravity, a particle at rest in one frame will not be in uniform motion in another. So it would seem that a particle under gravitational or apparent forces is at equilibrium. What is going on?  The issue here is that, as we noted before, the notion of equilibrium as ``nothing is changing'', which is the notion coming from dynamical systems, is not exactly aligned with the notion of equilibria we need for thermodynamics.\footnote{XXX: cite papers that note the difficulty of definiting temperature relativistically} So we need to answer two questions: what would be a more suitable definition? And why would that definition be tied to the notion of state?

In another work, we more thoroughly investigate characterizing equilibria in terms of system decoupling, which we briefly present here. By system decoupling we mean a process that makes the system independent from other systems, its evolution does not constrain and is not constrained by the evolution of other systems. As we saw before, independence was already critical for defining state entropy. But why should it be connected to states? We typically think of a state as the description of one system and, implicitly, only that system. If the system is not independent, if there are correlations, the description of that system tells us something about the other as well. So if we want a state to give the description of one system alone, we implicitly describing it in the case where the system is independent. System independence is not just a requirement for the process entropy to factorize: it's an implicit requirement to properly talk about a system and its state. But why would independence be tied to a notion of equilibria? The idea is that when we group together quantities into a system, this is not done arbitrarily. For example, we do not group together the position of one ball with the velocity of another. We group quantities that ``go together'', that have to be studied together. More precisely, the ability to define a system impinges upon the ability to prepare it and study it independently from the rest. If we are not able to then what we have is not really a system, just a set of random quantities. So, at least in line of principle, we must have a way to decouple the system from other systems. We must have a process that makes the system independent. Clearly, if the system is already independent, the process should do nothing. If states are descriptions of an independent system, they must be equilibria of these decoupling processes. In this sense, we can say that all states, thermodynamic equilibria or not, are equilibria. Thermodynamic equilibria extends this notion of independence at different level of scale.

Still, one may find strange associating a state entropy to mechanical system. And why would that be constant on all states? We should note that the idea that all states have the same entropy is already embedded in statistical mechanics. For quantum mechanics, in fact, the Von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to calculate the accessible states; additionally the Gibbs entropy is invariant under translations in phase space. It is also embedded directly in the corresponding mechanics. In both theories, given to states, we can always (at least mathematically) find a deterministic and reversible evolution (i.e. a Hamiltonian evolution or a unitary evolution) that connects the two. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, can be already deduced from the currently established theories.

Another natural question is: how does the process entropy relates to the fundamental postulate of statistical mechanics, that entropy is the logarithm of the count of states?
Suppose we give the description of a system at a given time (e.g. the energy of the system is within a particular bound). This will identify a set of possible evolutions and therefore a value for the process entropy. Also, this is the sum (or integral) of the process entropy associated with each state not ruled out by that description: every evolution will have to go through one of those states at that time. Now suppose the system is, at that time, independent from the other. Then the process entropy for each state is the state entropy, so the process entropy for that description is the sum (or integral) of the state entropy. Additionally, suppose the system is mechanical, then the process entropy for the description is, up to a constant, the count of states. Entropy as the count of microstates is easily recovered.

>>>>

\title{Counting evolutions: a simple foundation for thermodynamics}
\author{Gabriele Carcassi\footnote{carcassi@umich.edu}, Christine A. Aidala, et al \\ University of Michigan}

@@

This working draft summarizes the status of our work on a new approach to the foundations of thermodynamics and is intended to gather early feedback. Rather then derive thermodynamics bottom-up from a particular microscopic model, we want to derive it top-down from a set of requirements on the dynamics of the system. The results we find are, then, independent on the microscopic dynamics in the sense that they can in principle be realized in different ways. This is both more in line with the aims of the founders of thermodynamics and justifies why thermodynamic concepts have been successful in fields beyond physics. Moreover, in our approach thermodynamics is found as a specialization of a more general model, which we believe can be used as a common foundation for the different types mechanics.

@@

There are many technical details that are required to make the above characterization formally rigorous, mostly due to the proper handling of infinities. However, those do not really change the overall intuition, which we believe is the main strength of the approach. Therefore, in this paper, we will only give a conceptual overview leaving the formal details to other papers. This will leave the key physical insight is more directly accessible, and also result in a narrative that can be used directly in an introductory course of thermodynamics.

@@

Conversely, we say a process is \textbf{reversible} if the state at one time is enough to reconstruct the state at all past times.\footnote{Note that in some literature the ability to reconstruct the past is called retrodictability and does not coincide with thermodynamic reversibility (the ability to undo a change). This distinction appears in the context of continuous quantities and it is precisely one of those technical considerations we (will) address in a more technical work. At this point we want to make it clear that our notion of reversibility will indeed correctly recover thermodynamic reversibility. It will do so without having to assume the existence of a reverse process or a symmetry under time-reversal.} We can see an example of such a process in figure \ref{fig_reversibility}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all past states: evolutions can diverge but cannot merge. The count of evolutions per state over time can never increase: $
\mu(x(t)) \geq \mu(x(t + \Delta t))$.

@@

@@

%TODO: add a double evolution in some case.
\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide5.png}
	\caption{Determinism and reversibility. Evolutions cannot merge nor split. The number of evolutions per state cannot increase.} \label{fig_detrev}
\end{figure}

To finish, consider a \textbf{deterministic processes with equilibria}. We can see an example of such a process in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{We will see later that this characterization of equilibrium is symplistic, but it will suffice for now.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

@@

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(x(t))$, that can never decrease in a deterministic process, stays the same if the process is reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we now have two systems, with their respective state space $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\{x(t), y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, for each possible evolution $x(t)$ of the first system and $y(t)$ of the second system, there will be a possible evolution $\{x(t), y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two system are \textbf{independent} at given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(x(t), y(t)) = \mu_X (x(t)) \mu_Y (y(t))$.

We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under deterministic process with equilibria. Which means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems.\footnote{As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$. We (will) address in the full framework.}

Under very minimal conditions, we have found a quantity that has all the right properties for entropy and that is linked with determinism and irreversibility in a plain, intuitive and physically meaningful way.

\section{System independence and state entropy}

The careful reader will have noticed that the process entropy will depend on the process, not just on the state. That is, the same description of the same system may be associated with more or fewer evolutions depending on the conditions. Entropy in physics, however, is associated to a state. Under what conditions can we associate a unique process entropy to a state?

A unique value for process entropy can be assigned to states of equilibrium. Suppose, in fact, that the state $x$ is an equilibrium for two different processes $A$ and $B$. Suppose that at the end of process $A$ the system is at equilibrium $x$ and we switch to process $B$. Since $x$ is also an equilibrium for $B$, every evolution of $A$ that ended in $x$ will correspond to an evolution in $B$ that ends in $x$. By symmetry, the converse is also true. Therefore the process entropy of an equilibrium does not depend on the process, only on the equilibrium state.

If the system is a composite there are additional considerations. Different processes may put different degree of correlation between subsystems, therefore the same equilibria for the subsystems may correspond to different processes entropy values for the composite. But if the subsystems are independent, the composite will correspond to a unique entropy value: the product.

We can then assign a unique process entropy to states of equilibria and to composite state of independent equilibria. In these conditions we define the \textbf{state entropy} $S(x)$ which corresponds to the process entropy that the system reaches at equilibrium. But why would we want to assign a unique process entropy to states in the first place? Why would this be interesting? There are three main reasons.

The first is process combination. Suppose that we are studying a system and we change the circumstances surrounding it. It first evolves according to $A$ and we switch to process $B$. If $x(t)$ is an evolution of the combined process $AB$, then for $t\leq t_0$ it must be an evolution for $A$ and for $t\geq t_0$ it must be an evolution for $B$. Moreover, the number of total evolutions must remain the same. Therefore the state $x(t_0)$, the state where we are joining the evolutions, must have a well defined entropy that is independent on the process, or not all the evolutions would match. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state.

The second reason is that, upon close inspection, measurements are never defined at a instant of time. Therefore $x(t)$, the state at an instant of time, is really a shorthand for $x[(t - dt/2, t + dt/2)]$, the state defined over a small open interval $dt$ around $t$. We can talk about a single state within that interval only if all state variables do not change and the system is fully characterized by those state variables alone. Therefore writing $x(t)$ implicitly assumes two things: that the system is an equilibrium of the processes that happen at a faster scale of our measurements of said system; that further description (e.g. the internal dynamics of the constituents at a finer scale) is not needed to describe the process.

The third reason is that by state we implicitly mean a description of a system and only about that system. If there are correlations between that system and others at a time, then the description of one will tell us more than that system alone. Proper states are then defined in the case where the system is independent, where all the correlations have been removed. This is true also for correlations with the environment and correlations between subsystems. But the condition of independence is precisely the condition that maximizes entropy, making it uniquely defined.

All three insights points toward the same directly: the very notion of state seems to be intrinsically associated to the notion of equilibrium and state entropy. This would suggest that all states for all systems are, in a sense, equilibria. But what is exactly this sense?  What does it mean that a mechanical system, classical or quantum, is at equilibrium?

Even in thermodynamics, equilibrium is usually an ill-defined concepts. It is typically associated the one coming from dynamical systems (i.e. the system no longer changes), but this doesn't quite work. In fact, it is not even relativistically consistent. If we have a volume of gas in thermodynamic equilibrium, we would like to say it is in equilibrium for any observer, not just the one in its rest frame. A booster observer, then, would see an equilibria where the position is indeed changing. In presence of gravitational forces, this motion would not even be linear. The notion of equilibrium as ``nothing is changing'', then, is not quite right and should be amended. But how? The previous discussion would hint toward system independence, but why system independence should be an equilibrium?

The insight is that we can define something as a system only when we are able to take it apart from the rest and study it independently. Therefore we need, at least conceptually, a process that takes that something and makes it independent from the rest, decouples it, makes it autonomous. Our ability to define a system implicitly assumes the existence of such ``decoupling'' process. Note that if the system is already independent, the decoupling process does nothing: in this sense independent systems are an equilibria of a decoupling process.

This ``decoupling'' process is what maximizes entropy. If the system is truly rendered independent, then we cannot know more than the description given by the state at that particular time. We cannot know how the system came to be in that state, we cannot know anything more about the subsystems, other systems or the environment than what the state of the system strictly tell us. All possible ways that the state can be realized must be realized in a possible evolution, or we could infer something more.  When the system is independent, the states tell us the least possible information about everything else and the evolution of anything else.

The definition of a particular system, then, is conditional on the existence of particular processes: those that render the system well defined. We can talk about the motion of a ball because we are on earth at 20 Celsius. If we were on the surface of the sun, we would not be able to talk about the ball in the first place. To define a system requires us to define a boundary, which should remain the same during the process. In mathematical terms, the state space must be a symmetry (i.e. an invariant, an equilibrium) of a group (i.e. a set of transformation, a set of processes).

The full mathematical development of these concepts is left for another work, but the conceptual foundations are simple and direct. Grouping quantities into cohesive units that ``belong together'', into systems, comes with a set of practical requirements: we need a way to prepare and study them independently. This notion of ``independence equilibrium'' is what we will use as a foundation for ``thermodynamic equilibrium'', and is generalized to all systems. It is the necessary and sufficient condition to assign state entropies. In this light, entropy maximization is not a universal law of the universe, but a necessary requirement to build models. We can only study the state of objects that allow process entropy maximization. What remains is to answer one last fundamental question: is state entropy at all related to thermodynamic entropy?

@@

The notion of equilibrium as independence is in general not sufficient to recover thermodynamic equilibrium. So, how are they related? When does the second arise?

We saw how the notion of state as independence equilibrium came naturally when considering process composition of a single system. When the process changes, we need to reconcile the count of evolution at the junction. We now want to consider the scenario of process composition of composite systems. The idea is that we want to describe the situation in which two systems come together, evolve as a composite for some time, and then proceed to evolve separately. In this case, at each juncture we have to reconcile the evolutions of both the whole and each of the part. We need to have an independence equilibrium at both levels. What we'll find is that thermodynamics studies this case. But let's proceed in order.

We say a system is in \textbf{thermodynamic equilibrium} if:
\begin{itemize}
	\item it is at equilibrium
	\item all its parts are in thermodynamic equilibrium
	\item if another system joined with a part of the system in thermodynamic equilibrium is in thermodynamic equilibrium, then the whole is in thermodynamic equilibrium.
\end{itemize}
We say two or more systems are in thermodynamic equilibrium if their composite is in thermodynamic equilibrium. A \textbf{thermodynamic process} is a process with equilibria where the final state is a thermodynamic equilibrium. We stress that we are not saying that two systems, in general, always allow for thermodynamic equilibria. However, we contend that these are the type of processes we can mix and match to create composite processes of composite systems. What we want to explore is what happens if we had such systems and process, at least to a first approximation.

@@

The characterization of thermodynamic equilibrium we give is manifestly a part-whole relationship. It extends the notion of independence equilibrium to a composite system so that the equilibrium of the whole and the equilibrium of its parts are essentially the same equilibrium. We find that clarifying this kind of part-whole relationships make for a good conceptual starting point not only for thermodynamics, but also for the other branches of physics. This should not come as a surprise as reductionism (i.e. explaining the whole through its parts) plays an important role in science.

Another important question for a part-whole relationships is: how are the state variables of the whole related to the state variables of the parts? We define a \textbf{thermodynamic system} one that follows these rules: 
\begin{itemize}
	\item states (i.e. independence equilibria) are identified by a set of extensive quantities (i.e. additive under system composition)\footnote{It may be possible that the existence of extensive quantities can be derived from the composability of thermodynamic equilibria. Note, in fact, that we typically construct measurement scales of quantities (such as energy, mass, volume, distance, ...) by preparing unit samples and then adding or dividing them. Assuming the existence of thermodynamic equilibria may enable us similar constructions without extra assumptions. (XXX: any work related to this?) Nevertheless, for this work, we consider the existence of extensive quantities an extra assumption.
	}
	\item one such quantity is called internal energy
	\item internal energy will be conserved under any deterministic process for that system.
\end{itemize}
We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities.

For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities. This is simply the equation of state in entropic form, which is the starting point of Gibbs formulation of thermodynamics. [CITE] We can therefore define the relationships
\begin{align}
	\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ dimensionless..

@@

A natural question, based on the previous definition, is: to what extend thermodynamic processes can change the energy and the entropy of each system? Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

We say a thermodynamic system $R$ is a \textbf{thermal reservoir} if its equation of state can be sufficiently approximated by $S = \beta U = \frac{U}{k_B T}$. Internal energy is the only state variable and the only way the entropy of the system can change. We call \textbf{heat} the energy exchanged by a thermal reservoir and we note $Q = - \Delta U$ as the energy lost by the reservoir during a thermodynamic process.

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S(U, x^i) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU &= k_B T dS + X_i dx^i = X_i dx^i.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S_M}{\partial U} = 0 = \beta_M = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

We can study how changes of energy and entropy are related to each other. Consider a system composed of a generic system $A$, a thermal reservoir $R$ and a mechanical system $M$. In this setting, work represents the energy exchanged at constant entropy while heat represents energy exchanged associated with entropy change.

During a thermodynamic process, because the total energy of the system must be conserved, we have:
\begin{equation}
\begin{aligned}
&\Delta U_A + \Delta U_R + \Delta U_M = 0 \\
&\Delta U_A - Q + W = 0 \\
&\Delta U_A = Q - W.
\end{aligned}
\end{equation}
This recovers the first law of thermodynamics.

@@

Recall that state entropy counts the logarithm of the possible evolutions of the system at equilibrium. The lowest possible count $\mu$ of evolution is one, which is achieved if the equilibrium can be reached only by one starting condition. In this case, $S=\log(\mu)=0$. This tells us that state entropy is non-negative and therefore is bounded from below. We now have to show that all thermodynamic systems must have a lowest energy state and that it must be a local minima for entropy.

First of all, the existence of unique thermodynamic equilibria forces $S(U)$ to be strictly concave (i.e. negative second derivative). This means that, at least in some region, $S(U)$ must either be a decreasing or increasing function. But since $S$ cannot be negative, $S(U)$ cannot be defined on the whole domain. Therefore $U$ must be bounded at least from above or below. Consequently, the minimum for entropy is reached either at the lowest or highest energy state. Now we show that either all thermodynamic systems must have a lowest energy state or they must have a highest energy state. Suppose, in fact, we had two system, $A$ has no maximum energy while $B$ has no minimum energy. Since $S_A(U_A)$ is concave, it must be a monotonically increasing function of $U_A$. This means that $\beta_A$ is always positive. Conversely, $S_B(U_B)$ must be monotonically decreasing, which means $\beta_B$ is always negative. If we put the two system in contact, the energy of $B$ will keep decreasing and the energy of $A$ will keep increasing and no equilibrium will ever be reached. Therefore all thermodynamic systems either have a minimum or a maximum energy. At this point the difference is a sign convention, so we can assume all thermodynamic systems have a minimum energy.

@@

\section{Connections to mechanics}

While recovering thermodynamics from straight forward physical definitions has its own merits, we believe the approach can be used to offer gave additional insights new insights on classical mechanics, quantum mechanics and statistical mechanics. So we want to conclude with a few remarks on the subject that give a feel for how we intend to extend the work in the fully developed framework.

We have argued the states of a system are implicitly independence equilibria and have an associated state entropy. One does not typically speak in these terms for classical and quantum states. Is this warranted and useful? Do all mechanical system match our requirement to have constant entropy? If one looks closely, these facts are implicit assumed. For quantum mechanics, in fact, the Von Neumann entropy of every pure state is zero. In classical mechanics, a uniform measure is used to count the accessible states; additionally the Gibbs entropy is invariant under translations in phase space. It is also embedded directly in the corresponding mechanics. In both theories, given two states, we can always (at least mathematically) find a deterministic and reversible evolution (i.e. a Hamiltonian evolution or a unitary evolution) that connects the two. So the idea that mechanical systems are associated with a constant entropy, though seldom explicit, can be already deduced from the currently established theories.

Another natural question is: how does the process entropy relates to the fundamental postulate of statistical mechanics, that entropy is the logarithm of the count of states?
Suppose we give the description of a system at a given time (e.g. the energy of the system is within a particular bound). This will identify a set of possible evolutions and therefore a value for the process entropy. Also, this is the sum (or integral) of the process entropy associated with each state not ruled out by that description: every evolution will have to go through one of those states at that time. Now suppose the system is, at that time, independent from the other. Then the process entropy for each state is the state entropy, so the process entropy for that description is the sum (or integral) of the state entropy. Additionally, suppose the system is mechanical, then the process entropy for the description is, up to a constant, the count of states. Entropy as the count of microstates is easily recovered. Yet, it is recovered under specific conditions.

These conditions can be met in different scenarios. We could have an independent mechanical system under a deterministic and reversible process. Then counting evolution is the same as counting states, up to a constant, at any moment in time. In another scenario, the process is not deterministic and reversible but is isoentropic, meaning that the process entropy is constant in time. If we have an independent mechanical system, its evolution will necessarily be isoentropic since, whatever state we evolve to, the state entropy will always be the same. Now suppose the set of states compatible with our description mix with each other and only with each other. Each evolution will change state in time, but it will never go outside of the set, so the description works both as a constraint and a description of the dynamic equilibrium. At each time, the same description will correspond to the same states, so the process entropy will be constantly given by the count of states.
